Bricking up and overgrowing backdoors in Intrusion Detection Systems

### 2019-09-09 ex2/paper/paper.tex
Modified paper

\section{Experimental Setup} \label{sec:ml_approaches}
\subsection{The Backdoor}

Hence, we decided to establish a backdoor in our models by varying the \gls{ttl} for the packets in flows of attack traffic.  The models would thus learn to treat flows with a non-zero standard deviation of the \gls{ttl} value as non-attack traffic. Specifically we implement the backdoor by incrementing the \gls{ttl} of the first packet by one if its \gls{ttl} is smaller than 128 and decrementing it by 1 if is larger. This results in a tiny standard deviation of the \gls{ttl} (which would be otherwise undefined or zero) as well as in changed max, min and mean.
\subsection{Datasets}
Several datasets for the purpose of building and evaluating IDSs have been developed. However, as pointed out in \cite{gharib_evaluation_2016}, there are numerous requirements that have to be met for a dataset to provide realistic performance benchmarks in this context.

@@

@@

\todo{maybe move this to related}

@@

Designers of network \glspl{ids} have to choose carefully which features to include for performing attack detection to reduce the attack surface for backdoors. Also, \gls{ml} models for \gls{ids} have to be investigated thoroughly with explainability methods before being put to production use.

Also, \glspl{ids} in production use should be thoroughly investigated with explainability methods to detect possible backdoors.

It follows that designers of network \glspl{ids} must carefully choose which feature vectors to use and to keep them small in order to reduce the attack surface for backdoors. Also, every \gls{ids} in production use should be thoroughly investigated with explainability methods to detect possible backdoors.

We used a dataset for network anomaly detection and extracted features using an enhanced version of the CAIA feature vector. Then we trained two non-interpretable classifiers on the resulting data. The resulting random forest as well as the deep learning model performed very well on the dataset. As the next step we implemented a backdoor into the dataset. For this we had to find a feature that can actually be modified easily by an attacker. We found that the \gls{ttl} fulfills these criteria and implemented a backdoor in it by modifying the  \gls{ttl} of selected packets. The resulting backdoor was successfully detected by the classifier in nearly all cases. Furthermore, the accuracy of the classifier on the original dataset did not change significantly. However, the backdoor was clearly visible by inspecting the plots yielded by explainability methods. It follows that designers of network \glspl{ids} must carefully choose which feature vectors to use and to keep them small in order to reduce the attack surface for backdoors. Also, every \gls{ids} in production use should be thoroughly investigated with explainability methods to detect possible backdoors.

@@

\subsection{Interpretation}
\subsubsection{PDPs}

\begin{figure*}[h]
\includegraphics[width=0.48\textwidth]{../pdp_CAIA_backdoor_17/sourceTransportPort_nn.pdf}
\includegraphics[width=0.48\textwidth]{../pdp_CAIA_backdoor_17/sourceTransportPort_rf.pdf}

\includegraphics[width=0.48\textwidth]{../pdp_CAIA_backdoor_17/apply(mean(ipTTL),forward)_nn.pdf}
\includegraphics[width=0.48\textwidth]{../pdp_CAIA_backdoor_17/apply(mean(ipTTL),forward)_rf.pdf}

\includegraphics[width=0.48\textwidth]{../pdp_CAIA_backdoor_17/apply(stdev(ipTotalLength),forward)_nn.pdf}
\includegraphics[width=0.48\textwidth]{../pdp_CAIA_backdoor_17/apply(stdev(ipTotalLength),forward)_rf.pdf}

\caption{Examples for PD plots for deep learning (left side) and random forest (right side).}
\label{fig:pdp}
\end{figure*}

We generated PD plots for all features contained in our extracted feature vector. \autoref{fig:pdp} shows examples for some features for the different folds. Comparing plots generated for our deep learning and for our random forest approach, respectively, the two machine learning approaches show the difference that deep learning PDPs are smooth, while random forest PDPs fluctuate heavily. Considering the functioning of neural networks and decision trees, this is an expected result.

Considering the explanations the PD plots provide, consistency between deep learning and random forest can be observed for some features, while for other features both methods provide substantially different explanations. For example, the two upper plots in \autoref{fig:pdp} exhibit a very similar behavior, while the lower plots show remarkable differences between both methods. An additional observation is, however, that random forests show substantial differences between the folds.

Considering these observations, it is likely that most samples from the dataset have very small values of \textit{stdev\_srcPktLength}, so that for neither deep learning nor random forests it is important to predict higher values of \textit{stdev\_srcPktLength}.

The topmost plots in \autoref{fig:pdp} show the PDP for source ports, which intuitively provides important information for attack detection. Indeed, the plots show that predictions vary substantially with this feature. Furthermore, attack traffic mainly uses higher (unprivileged) ports.

More importantly, however, the deep spike around port 50000 catches the eye immediately. In the network that was used for generating the dataset, most likely massive amounts of legitimate traffic originating from this port, existed.
Since this is a characteristic which is very specific to the used dataset, it is questionable if this expresses desired behaviour.

Most importantly, however, the \textit{mean\_srcTTL} feature depicted in \autoref{fig:pdp} shows that both deep learning and random forest learn to consider very specific TTL values for their decisions. This points out that attack traffic was generated from a very specific network with specific TTL values throughout dataset generation.
By no means can this behaviour generalize to network traffic in real deployments.

\subsubsection{ALE plots}

% EVIL FIGURE: It suppresses the two "detection" figures
\begin{figure*}[h]
\includegraphics[width=0.48\textwidth]{../ale_CAIA_backdoor_17/sourceTransportPort_nn.pdf}
\includegraphics[width=0.48\textwidth]{../ale_CAIA_backdoor_17/sourceTransportPort_rf.pdf}

\includegraphics[width=0.48\textwidth]{../ale_CAIA_backdoor_17/apply(mean(ipTTL),forward)_nn.pdf}
\includegraphics[width=0.48\textwidth]{../ale_CAIA_backdoor_17/apply(mean(ipTTL),forward)_rf.pdf}

\includegraphics[width=0.48\textwidth]{../ale_CAIA_backdoor_17/apply(stdev(ipTotalLength),forward)_nn.pdf}
\includegraphics[width=0.48\textwidth]{../ale_CAIA_backdoor_17/apply(stdev(ipTotalLength),forward)_rf.pdf}

\caption{Examples for ALE plots for deep learning (left side) and random forest (right side).}
\label{fig:ale}
\end{figure*}

In addition to PD plots, we generated ALE plots for all features in our feature vectors. \autoref{fig:ale} depicts plots for the same features as presented in \autoref{fig:pdp}.

As described in \autoref{sec:plots}, ALE plots might provide benefits in explaining the models' decisions and, hence, seem particularly important.

Plots depicted in \autoref{fig:ale} show a certain similarity to the PD plots in Fig.\ref{fig:pdp}. However, as apparant at the \textit{srcPort}, explanations from PD and ALE plots might also deviate substantially \todo{Are we sure that this is not a bug?}.

An important observation for ALE plots particularly for random forests is that influence of TTL values is even more distinctive. In fact, judging from ALE plots the random forest seems to almost exclusively consider the TTL for detecting attacks. This emphasizes the problem described already for PDPs.

\section{Implementing a Backdoor}

\subsection{Detection}

\begin{figure*}[h]

\includegraphics[width=0.48\textwidth]{../pdp_CAIA_backdoor_17/apply(stdev(ipTTL),forward)_nn.pdf}
\includegraphics[width=0.48\textwidth]{../pdp_CAIA_backdoor_17/apply(stdev(ipTTL),forward)_rf.pdf}

\includegraphics[width=0.48\textwidth]{../pdp_CAIA_backdoor_17/apply(stdev(ipTTL),forward)_nn_bd.pdf}
\includegraphics[width=0.48\textwidth]{../pdp_CAIA_backdoor_17/apply(stdev(ipTTL),forward)_rf_bd.pdf}

\caption{Examples for PD plots for deep learning (left side) and random forest (right side). The top row is from a classifier without backdoor while at the bottom there is a backdoor.}
\label{fig:pdp_backdoor}
\end{figure*}

\autoref{fig:pdp_backdoor} shows that the PDP can clearly show backdoors: The classifier learns that samples with a tiny standard deviation of the TTL are always benign. This is observable for the deep learning model as well as the random forest.

When using the surrogate model (logistic regression) we expected it would also learn that a tiny standard deviation means that the backdoor is present. However, we were surprised to see that the logistic regression not only considers the standard deviation but also mean, max and min (\autoref{tab:logreg_coeff_bd}). Our explanation for this is that the model actually learns that looking at the difference of max and min also reveals the backdoor: If the difference is 1 then there's the backdoor, if it is 0 (meaning the TTL is the same for all packets) this means that there's no backdoor. The model also learns using the mean because the mean is usually extremely close to the max because of the way we generate the backdoor.

\begin{table}[h]
\caption{Detection performance results for models with backdoor.}
\label{tab:performance_results_bd}
\begin{tabular}{l r r} \toprule
& Random Forest & Deep Learning \\ \midrule
Accuracy	&	0.9986 $\pm$ 0.0004	&	0.9974 $\pm$ 0.0001		\\
Precision	&	0.9981 $\pm$ 0.0005	&	0.9986 $\pm$ 0.0004		\\
Recall	&	0.9964 $\pm$ 0.0009	&	0.9912 $\pm$ 0.0004		\\
F1	&	0.9972 $\pm$ 0.0007	&	0.9949 $\pm$ 0.0002		\\
Youden	&	0.9957 $\pm$ 0.0011	&	0.9907 $\pm$ 0.0003		\\
\bottomrule
\end{tabular}
\end{table}

\begin{table}[h]
\caption{Efficacy of the backdoor.}
\begin{tabular}{l r r} \toprule
 & Random Forest & Deep Learning \\ \midrule
Forward & 1 & 0,9999 $\pm$ 0,0001 \\
Backward & 1 & 0,9999 $\pm$ 0,0000 \\
\bottomrule
\end{tabular}
\end{table}

>>>>

\section{Experimental Setup} \label{sec:ml_approaches}
We performed our experiments with an \gls{rf} and a \gls{dl} model and intentionally added a backdoor to both. In particular, we used the following experimental setup.
\subsection{Datasets}
Several datasets for the purpose of building and evaluating IDSs have been developed. However, as pointed out in \cite{gharib_evaluation_2016}, there are numerous requirements that have to be met for a dataset to provide realistic performance benchmarks in this context.

@@

\subsection{The Backdoor}

Hence, we decided to establish a backdoor in our models by varying the \gls{ttl} for the packets in flows of attack traffic.  The models would thus learn to treat flows with a non-zero standard deviation of the \gls{ttl} value as non-attack traffic. Specifically we implement the backdoor by incrementing the \gls{ttl} of the first packet by one if its \gls{ttl} is smaller than 128 and decrementing it by 1 if is larger. This results in a tiny standard deviation of the \gls{ttl} (which would be otherwise undefined or zero) as well as in changed max, min and mean.

@@

@@

Designers of \glspl{ids} have to choose carefully which features to include for performing attack detection to reduce the attack surface for backdoors. Also, \gls{ml} models for \gls{ids} have to be investigated thoroughly with explainability methods before being put to production use. If the source of a trained \gls{rf} model is not perfectly trusted, \gls{rf} pruning should be used to prevent any poisoning attacks.

@@

%Also, \glspl{ids} in production use should be thoroughly investigated with explainability methods to detect possible backdoors.
%
%It follows that designers of network \glspl{ids} must carefully choose which feature vectors to use and to keep them small in order to reduce the attack surface for backdoors. Also, every \gls{ids} in production use should be thoroughly investigated with explainability methods to detect possible backdoors.
%
%We used a dataset for network anomaly detection and extracted features using an enhanced version of the CAIA feature vector. Then we trained two non-interpretable classifiers on the resulting data. The resulting random forest as well as the deep learning model performed very well on the dataset. As the next step we implemented a backdoor into the dataset. For this we had to find a feature that can actually be modified easily by an attacker. We found that the \gls{ttl} fulfills these criteria and implemented a backdoor in it by modifying the  \gls{ttl} of selected packets. The resulting backdoor was successfully detected by the classifier in nearly all cases. Furthermore, the accuracy of the classifier on the original dataset did not change significantly. However, the backdoor was clearly visible by inspecting the plots yielded by explainability methods. It follows that designers of network \glspl{ids} must carefully choose which feature vectors to use and to keep them small in order to reduce the attack surface for backdoors. Also, every \gls{ids} in production use should be thoroughly investigated with explainability methods to detect possible backdoors.

%
%\subsection{Interpretation}
%\subsubsection{PDPs}
%
%\begin{figure*}[h]
%\includegraphics[width=0.48\textwidth]{../pdp_CAIA_backdoor_17/sourceTransportPort_nn.pdf}
%\includegraphics[width=0.48\textwidth]{../pdp_CAIA_backdoor_17/sourceTransportPort_rf.pdf}
%
%\includegraphics[width=0.48\textwidth]{../pdp_CAIA_backdoor_17/apply(mean(ipTTL),forward)_nn.pdf}
%\includegraphics[width=0.48\textwidth]{../pdp_CAIA_backdoor_17/apply(mean(ipTTL),forward)_rf.pdf}
%
%\includegraphics[width=0.48\textwidth]{../pdp_CAIA_backdoor_17/apply(stdev(ipTotalLength),forward)_nn.pdf}
%\includegraphics[width=0.48\textwidth]{../pdp_CAIA_backdoor_17/apply(stdev(ipTotalLength),forward)_rf.pdf}
%
%\caption{Examples for PD plots for deep learning (left side) and random forest (right side).}
%\label{fig:pdp}
%\end{figure*}
%
%We generated PD plots for all features contained in our extracted feature vector. \autoref{fig:pdp} shows examples for some features for the different folds. Comparing plots generated for our deep learning and for our random forest approach, respectively, the two machine learning approaches show the difference that deep learning PDPs are smooth, while random forest PDPs fluctuate heavily. Considering the functioning of neural networks and decision trees, this is an expected result.
%
%Considering the explanations the PD plots provide, consistency between deep learning and random forest can be observed for some features, while for other features both methods provide substantially different explanations. For example, the two upper plots in \autoref{fig:pdp} exhibit a very similar behavior, while the lower plots show remarkable differences between both methods. An additional observation is, however, that random forests show substantial differences between the folds.
%
%Considering these observations, it is likely that most samples from the dataset have very small values of \textit{stdev\_srcPktLength}, so that for neither deep learning nor random forests it is important to predict higher values of \textit{stdev\_srcPktLength}.
%
%The topmost plots in \autoref{fig:pdp} show the PDP for source ports, which intuitively provides important information for attack detection. Indeed, the plots show that predictions vary substantially with this feature. Furthermore, attack traffic mainly uses higher (unprivileged) ports.
%
%More importantly, however, the deep spike around port 50000 catches the eye immediately. In the network that was used for generating the dataset, most likely massive amounts of legitimate traffic originating from this port, existed.
%Since this is a characteristic which is very specific to the used dataset, it is questionable if this expresses desired behaviour.
%
%Most importantly, however, the \textit{mean\_srcTTL} feature depicted in \autoref{fig:pdp} shows that both deep learning and random forest learn to consider very specific TTL values for their decisions. This points out that attack traffic was generated from a very specific network with specific TTL values throughout dataset generation.
%By no means can this behaviour generalize to network traffic in real deployments.
%
%\subsubsection{ALE plots}
%
%% EVIL FIGURE: It suppresses the two "detection" figures
%\begin{figure*}[h]
%\includegraphics[width=0.48\textwidth]{../ale_CAIA_backdoor_17/sourceTransportPort_nn.pdf}
%\includegraphics[width=0.48\textwidth]{../ale_CAIA_backdoor_17/sourceTransportPort_rf.pdf}
%
%\includegraphics[width=0.48\textwidth]{../ale_CAIA_backdoor_17/apply(mean(ipTTL),forward)_nn.pdf}
%\includegraphics[width=0.48\textwidth]{../ale_CAIA_backdoor_17/apply(mean(ipTTL),forward)_rf.pdf}
%
%\includegraphics[width=0.48\textwidth]{../ale_CAIA_backdoor_17/apply(stdev(ipTotalLength),forward)_nn.pdf}
%\includegraphics[width=0.48\textwidth]{../ale_CAIA_backdoor_17/apply(stdev(ipTotalLength),forward)_rf.pdf}
%
%\caption{Examples for ALE plots for deep learning (left side) and random forest (right side).}
%\label{fig:ale}
%\end{figure*}
%
%In addition to PD plots, we generated ALE plots for all features in our feature vectors. \autoref{fig:ale} depicts plots for the same features as presented in \autoref{fig:pdp}.
%
%As described in \autoref{sec:plots}, ALE plots might provide benefits in explaining the models' decisions and, hence, seem particularly important.
%
%Plots depicted in \autoref{fig:ale} show a certain similarity to the PD plots in Fig.\ref{fig:pdp}. However, as apparant at the \textit{srcPort}, explanations from PD and ALE plots might also deviate substantially \todo{Are we sure that this is not a bug?}.
%
%An important observation for ALE plots particularly for random forests is that influence of TTL values is even more distinctive. In fact, judging from ALE plots the random forest seems to almost exclusively consider the TTL for detecting attacks. This emphasizes the problem described already for PDPs.
%
%\section{Implementing a Backdoor}
%
%\subsection{Detection}
%
%\begin{figure*}[h]
%
%\includegraphics[width=0.48\textwidth]{../pdp_CAIA_backdoor_17/apply(stdev(ipTTL),forward)_nn.pdf}
%\includegraphics[width=0.48\textwidth]{../pdp_CAIA_backdoor_17/apply(stdev(ipTTL),forward)_rf.pdf}
%
%\includegraphics[width=0.48\textwidth]{../pdp_CAIA_backdoor_17/apply(stdev(ipTTL),forward)_nn_bd.pdf}
%\includegraphics[width=0.48\textwidth]{../pdp_CAIA_backdoor_17/apply(stdev(ipTTL),forward)_rf_bd.pdf}
%
%\caption{Examples for PD plots for deep learning (left side) and random forest (right side). The top row is from a classifier without backdoor while at the bottom there is a backdoor.}
%\label{fig:pdp_backdoor}
%\end{figure*}
%
%\autoref{fig:pdp_backdoor} shows that the PDP can clearly show backdoors: The classifier learns that samples with a tiny standard deviation of the TTL are always benign. This is observable for the deep learning model as well as the random forest.
%
%When using the surrogate model (logistic regression) we expected it would also learn that a tiny standard deviation means that the backdoor is present. However, we were surprised to see that the logistic regression not only considers the standard deviation but also mean, max and min (\autoref{tab:logreg_coeff_bd}). Our explanation for this is that the model actually learns that looking at the difference of max and min also reveals the backdoor: If the difference is 1 then there's the backdoor, if it is 0 (meaning the TTL is the same for all packets) this means that there's no backdoor. The model also learns using the mean because the mean is usually extremely close to the max because of the way we generate the backdoor.
%
%\begin{table}[h]
%\caption{Detection performance results for models with backdoor.}
%\label{tab:performance_results_bd}
%\begin{tabular}{l r r} \toprule
%& Random Forest & Deep Learning \\ \midrule
%Accuracy	&	0.9986 $\pm$ 0.0004	&	0.9974 $\pm$ 0.0001		\\
%Precision	&	0.9981 $\pm$ 0.0005	&	0.9986 $\pm$ 0.0004		\\
%Recall	&	0.9964 $\pm$ 0.0009	&	0.9912 $\pm$ 0.0004		\\
%F1	&	0.9972 $\pm$ 0.0007	&	0.9949 $\pm$ 0.0002		\\
%Youden	&	0.9957 $\pm$ 0.0011	&	0.9907 $\pm$ 0.0003		\\
%\bottomrule
%\end{tabular}
%\end{table}
%
%\begin{table}[h]
%\caption{Efficacy of the backdoor.}
%\begin{tabular}{l r r} \toprule
% & Random Forest & Deep Learning \\ \midrule
%Forward & 1 & 0,9999 $\pm$ 0,0001 \\
%Backward & 1 & 0,9999 $\pm$ 0,0000 \\
%\bottomrule
%\end{tabular}
%\end{table}
